---
title: "Setting Up R for Bioinformatics Workflows"
description: |
    A Windows user oriented guide to the initial set-up for R.
author:
    - name: Feifei Li
      url: {}
date: 6-03-2021
bibliography: biblio.bib
output:
    distill::distill_article:
      self_contained: false
      toc: TRUE
categories:
    - R
    - bioconductor
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

I decided to write this because some of the steps in setting up R for
the new version of Bioconductor could get painful on Windows,
and they never teach you this in school.
On UNIX/Linux getting things set up could simply take one or two commands.

## R Setup

To readers who would like to follow the guide to set up R step-by-step,
I recommend you set up the package directory first
(see [R Package Directory](#r-pkg-dir)) if you have never done so.

### R Version

We are going to use Bioconductor extensively in this workflow.
Each version of Bioconductor has different requirements for R version
(see [Bioconductor](https://www.bioconductor.org/install/)).
Here I decide to use the latest version of Bioconductor
(till the date this post is created),
which requires R version __4.1.0__ or higher.

To check if you have the right version of R installed:

```{r updateR, eval=FALSE}
## Get the current version of R on your computer
Rver <- unlist(
    regmatches(R.version.string,
               regexec("(\\d+).(\\d+).(\\d+)", R.version.string)
    )
)[1]
## Update R if your R version is below the requirement
if (Rver <= "4.1.0") {
    if (!requireNamespace("installr", quietly = TRUE)) {
        if(!requireNamespace("devtools", quietly = TRUE)) {
            ## install installr from CRAN if no devtools
            install.packages("installr")
        } else {
            ## use devtools to install the latest installr version from GitHub
            devtools::install_github('talgalili/installr')
        }
    }
    installr::updateR()
}
```

<code>requireNamespace("installr")</code> checks whether the package 
<code>installr</code> was already installed in R before.
It returns a boolean value indicating whether the package was installed,
along with a message prompted in R console
if you didn't set <code>quitely = TRUE</code>.
You might have seen many people uses <code>require()</code>,
which also returns a boolean value.
The difference is, <code>require</code> also automatically loads
the package if installed, or downloads if not.
This means using <code>requireNamespace("installr")</code> is always
a better option for checking package installation.
You should always avoid <code>load()</code>.


I didn't load the package with <code>library(package.name)</code>
to use the function in the package, which is commonly what people do.
Instead, I call the function by <code>package.name::function.name()</code>,
and I encourage the readers to do this
because it explicitly tells those people who are reading your code
from which packages those functions come from.
It might also help you understand better when you come back to look at your code.
I prefer to keep things explicitly because it avoids confusions.
However, there indeed exists exceptions where a package is required to be loaded
using <code>library()</code>,
which we will see later when we try to map HGNC symbols from Entrez ID's.

Enough of lecturing, we come back to the R setup itself.
Once the code above is run, you will be prompted to install the latest version of R
if the current version of R on your computer does not meet the requirement.
Once the installation is complete,
restart RStudio.
Don't just use the <code>Session > Restart R</code>
because it won't switch to the latest installed R version in RStudio.
In the new R session,
check your package library directory with <code>.libpaths()</code> as
the newly updated R will change it to the R version specific directory.
If the User installation was selected during the installation,
the directory could be <code>C:\User\$env:USERNAME\Documents\R\win-library</code>.
If the default system installation was selected, it could be
<code>C:\Program Files\R\R-4.1.0\library</code>.
Sometimes it ends up in
<code>X:\{your R directory}\R-4.1.0\library</code>.

You might want to change it if these are not what you want.

### R Package Directory {#r-pkg-dir}

For me, I have a dedicated directory for R packages.
This way, I won't need to suffer from re-installing or migrating R packages
from the previous version of R after an update.
To change the default R package directory, enter the following in PowerShell:

```{powershell}
Add-Content C:\Users\$env:USERNAME\Documents\.Renviron R_LIBS="{path to your package directory}"
Add-Content C:\Users\$env:USERNAME\Documents\.Renviron R_LIBS_USER="{path to your package directory}\\user"
Add-Content C:\Users\$env:USERNAME\Documents\.Renviron R_LIBS_SITE="{path to your package directory}\\site"
```

Don't forget double slashes in the directory path.

If this is your first time setting up a default package library,
after the update you will have to re-install <code>knitr</code> and <code>rmarkdown</code>.


## Bioconductor

A must-have for bioinformatics workflows.

### BiocManager

Here we will install the latest release of Bioconductor 3.13:

```{r install-Bioc, warning=FALSE, echo=FALSE, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.13")
```

It doesn't get you the whole Bioconductor installed.
This is because Bioconductor itself is not really "a piece of software",
but a collection of over 1000 R packages,[@bioc]
which takes forever to install on a single machine at a time.
So <code>BiocManager</code> is more of a package manager like the CRAN project,
just as its package name suggests.

### Packages

Bioconductor follows a package release schedule different from CRAN,
so to install Bioinformatics, we don't use <code>install.packages()</code>,
but instead, for example, to install <code>edgeR</code>,
a package extensively used in processing RNA-seq data:

```{r install-edgeR, warning=FALSE, echo=FALSE, eval=FALSE}
if (!requireNamespace("edgeR", quietly = TRUE))
    BiocManager::install("edgeR", ask = FALSE)
```

## R Startup Behaviour

<code>Rprofile.site</code> file allows users to define the behaviours
at starup and the end of an R session.
R will source <code>Rprofile.site</code> at startup.
On Windows, it is located in <code>X:\{path to your R installation}\R-4.1.0\etc</code>.
It should be there. If not, just create an empty text file and rename it.
In case you confuse it with <code>.Rprofile</code>,
they are equivalent; the latter is the way they name it on UNIX/Linux.

### Tab Width

RStudio by deafult uses a tab width of 2 spaces,
which is consistent with the Google’s R Style Guide and the Tidyverse Style Guide.
If you would like to follow the Bioconductor Style Guide, which uses 4 spaces, then
add this to your <code>Rprofile.site</code> at the top:

```{r}
options(tab.width = 4)
```

And welcome to the 4-tab camp:)

### .First

<code>.First</code> is the function in <code>Rprofile.site</code> that
actually allows you to define the startup behaviour of an R session:

```{r}
.First <- function(){
    the first thing you want at startup
    the second thing you want at startup
}
```


#### Working Directory

R resets to a default "working directory" (most likely where you installed R)
every time you open it,
if your RStudio is not loaded with a R project <code>.Rproj</code>.
It could get annoying that you have to set it to your own the working directory mannualy 
(if you have created one) with

```{r setwd, eval=FALSE}
.First <- function(){
    setwd("{path to your working directory}")
}
```

You can also load your helper functions from your utility scripts at startup:

```{r load-util, eval=FALSE}
.First <- function(){
    setwd("{path to your working directory}")
    source("{path to your utility scripts}")
}
```

### .Last

Similar to <code>.First</code>,
it defines the end behaviour of an R session.

#### Saving R Objects

To save a single R object (variables, functions, dataframe, etc.) as a file
everytime closing an R session:

```{r}
.Last <- function(){
    save(object_to_save, file = ".\\data\\name_of_object.RData")
}
```

where <code>data</code> is a directory to store data in the working directory,
if you have created one.

And to load the object:

```{r}
load(".\\data\\name_of_object.RData")
```

which could also be added to your <code>.First</code>
if you need to use the object every time.

## Writing at the end

I know _Docker_ is a popular option out there for bioinformatics pipelines.
It creates a consistent environment specific to the pipeline
regardless of the operating system it is running on,
and saves this hustle to configure R.
The out-of-box feature is nice, but the downside is,
Docker containers running in the background
eats up computational resources (e.g. memory).
In addition, if R runs in a Docker container,
when communicating with some bioinformatics tools running on the host via their R API,
some re-mapping of files or ports will be required.
